---
title: "EDS 223 HW 1: Map making practice"
author: "Emalia Partlow"
warning: false
---

## Load packages
```{r}
library(tidyverse)
library(sf)
library(here)
library(tmap)
library(janitor)
library(kableExtra)
library(dplyr)
```

## Load data
```{r}
# Read in geodatabase of EJScreen data at the Census Block Group level
ejscreen <- sf::st_read(here::here("data", "ejscreen","EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb")) |> 
    clean_names()
```

## Initial data exploration
```{r}
# Check dataframe dimensions
dim(ejscreen)

# Look at first 3 rows
head(ejscreen, n = 3) |> 
  kable()|> 
  kable_styling()
```

## Filter to Monterey county
```{r}
# Filter to Monterey
monterey <- ejscreen |> 
    dplyr::filter(st_abbrev == 'CA') |> 
    dplyr::filter((cnty_name == 'Monterey County'))

# Check new dimensions
dim(monterey)

# Check that filtered to Monterey correctly
monterey |> 
    distinct(cnty_name)
```

## Map of people of color in Monterey
```{r}
tm_shape(monterey) +
tm_graticules() +
    tm_polygons(
        fill = 'p_peopcolorpct',
        fill.scale = tm_scale_continuous(values = c('white', 'darkgreen')),
        fill.legend = tm_legend(title = 'Percent people of color',
                            position = tm_pos_out('right', 'center')) 
    ) +
tm_compass(position = c('right', 'top')) +
tm_scalebar(position = c('left', 'bottom')) +
tm_title('People of color in Monterey') +
tm_legend(position = tm_pos_out('right', 'center'))
```

## Map of Superfund proximity in Monterey
```{r}
tm_shape(monterey) +
tm_graticules() +
    tm_polygons(
        fill = 'p_pnpl',
        fill.scale = tm_scale_continuous(values = c('white', 'darkgreen')),
        fill.legend = tm_legend(title = 'Superfund proximity percentile',
                            position = tm_pos_out('right', 'center')) 
    ) +
tm_compass(position = c('right', 'top')) +
tm_scalebar(position = c('left', 'bottom')) +
tm_title('Superfund proximity in Monterey') +
tm_legend(position = tm_pos_out('right', 'center'))
```

## Map interpretation

The percent proximity to a Superfund site is highest at the northern end of Monterey county, which is also where there is the highest percentage of people of color. The areas closest to a Superfund site are consistent with the location of Fort Ord, a former U.S Army base that was established as a Superfund site by the EPA in 1990. These maps display that people of color may be more exposed to Fort Ord's contaminated soil, groundwater, and unexploded ordinances due to their close proximity to the Superfund site.